D.I.A.G.R.A.M.: Development of Image Analysis for Graph Recognition And Modeling

### 2025-06-21 doc/article/article.tex

\subsubsection{Model}
TODO \\

@@

\subsection{Classifier}
The first architecture explored for the Classifier module was a Convolutional Neural Network (CNN) following AlexNet's architecture, but smaller in size.
It had two Convolutional layers, each followed by a batch normalization and then by a max pooling layers. The first layer had only 4 filters, while the second one had 8.
The output of the second layer was then flattened and passed through a fully connected layer which downsized the features to 128, followed by a ReLU activation function. Other two fully connected layer brought the number of output features down to the number of classes.
Then, a softmax activation function was applied to the output layer, in order to obtain the probabilities of each class.
The model was trained on our own dataset, with only three possible output classes: 'flowchart', 'graph' and 'other'; it was trained for 10 epochs, with an AdamW optimizer (with a learning rate of 1e-3) and a standard CrossEntropy Loss function.
\\

@@

\subsection{Arrow Orientation}
As we've seen in the previous section, the need to find the exact position of an arrow's head arises (see \textbf{Appendix A} for a full overview of the first approach).

@@

\section{Double Clustering}
The first algorithmic approach proposed to solve this task - called \textit{Double Clustering} - was as follows: 
\begin{enumerate}
	\item Apply a clustering algorithm to the bounding box.
	\item Suppose that the biggest cluster belongs to the arrow; the other clusters are thought to contain useless elements for this task, and can be discarded.
	\item Find the keypoints of the arrow using SIFT.
	\item Find the two keypoints inside the arrow with maximum distance. Assume that these two points are closely related to the two ends of the arrow.
	\item Run a template matching procedure inside the two clusters to find the head of the arrow.
\end{enumerate}

@@

>>>>

\subsubsection{Model}
Since it was assumed that classifying two types of diagrams - with a third 'other' class - was a quite simple task, a small network was deemed enough. The model consists of three convolutional blocks - each containing a convolutional layer, followed by a batch normalization and a max-pooling layer - then followed by three linear layers (for a more detailed overview, see Appendix~\ref{classification_net}). The output vector of the MLP contains three features - one for each class that this specific classifier needs to address. In order to determine the prediction, the softmax function is then applied. As later seen in section~\ref{exp:classifier}, a weighted Cross-Entropy loss is then used to train the model.  \\

@@

\subsection{Classifier}
\label{exp:classifier}
The first architecture explored for the Classifier module was a Convolutional Neural Network (CNN) following AlexNet's architecture, but smaller in size.
It had two Convolutional layers, each followed by a batch normalization and then by a max pooling layers. The first layer had only 4 filters, while the second one had 8.
The output of the second layer was then flattened and passed through a fully connected layer which downsized the features to 128, followed by a ReLU activation function. Other two fully connected layer brought the number of output features down to the number of classes.
Then, a softmax activation function was applied to the output layer, in order to obtain the probabilities of each class.
The model was trained on our own dataset, with only three possible output classes: 'flowchart', 'graph' and 'other'; it was trained for 10 epochs, with an AdamW optimizer (with a learning rate of 1e-3) and a standard CrossEntropy Loss function.
\\

@@

\subsection{Arrow Orientation}
As we've seen in the previous section, the need to find the exact position of an arrow's head arises (see \textbf{Appendix~\ref{double_clustering}} for a full overview of the first approach).

@@

\section{Double Clustering}
\label{double_clustering}
The first algorithmic approach proposed to solve this task - called \textit{Double Clustering} - was as follows: 
\begin{enumerate}
	\item Apply a clustering algorithm to the bounding box.
	\item Suppose that the biggest cluster belongs to the arrow; the other clusters are thought to contain useless elements for this task, and can be discarded.
	\item Find the keypoints of the arrow using SIFT.
	\item Find the two keypoints inside the arrow with maximum distance. Assume that these two points are closely related to the two ends of the arrow.
	\item Run a template matching procedure inside the two clusters to find the head of the arrow.
\end{enumerate}

@@

\section{The Classification Network}
\label{classification_net}

\begin{table}[h]
	\centering
	\caption{Classification Network Details}
	\label{tab:classification-architecture}
	\begin{tabular}{|c|c|c|c|}
		\hline
		\textbf{Conv Layer} & \textbf{Output Channels} & \textbf{Kernel Size} & \textbf{Pool Kernel Size} \\
		\hline 
		1 & 8 & 2 & 3 \\
		2 & 16 & 2  & 3 \\
		3 & 24 & 3  & 2 \\
		\hline
		\textbf{Linear Layer} & \textbf{Output Features} & \textbf{Droput Rate} & \textbf{Activation Function} \\
		\hline 
		1 & 2048 & 0.2 & ReLU \\
		2 & 128 & 0.2 & ReLU \\
		3 & 3 & 0.2 & ReLU \\
		\hline
	\end{tabular}
\end{table}
